
\mychapter{Abelian Integrals: The Logarithmic Components}

{\bf THIS CHAPTER IS VERY VAGUE AND INCOMPLETE.}

The techniques of the previous chapter suffice to compute the {\it
algebraic} portion of an Abelian integral, which is to say, an
algebraic function.  Liouville's theorem, however, tells us that there
can also be logarithmic components in the integral.  The two can be
easily separated, since the rational portion of the integral
corresponds to poles of second order poles and higher in the integrand,
while the logarithmic portion corresponds to first order poles
in the integrand.


\vfill\eject
\mysection{Examples}

\example Compute $\int {1\over{\sqrt{1-x^2}}} \,dx$

This is a familiar example from first year calculus, but let's
approach it using the techniques of this book.  The obvious attempt is
to use the algebraic extension $y^2=1-x^2$ and integrate ${1\over
y}\,dx$.

But we first need to determine if this differential has any poles at
infinity.  Calling {\tt puiseux} with {\tt deg=-1} computes just
the principle part of the differential:

\begin{maximablock}
puiseux(y^2 - (1 - x^2), x, y, inf, false,
        -1, 1/y * del(x));
\end{maximablock}

Since these expansions are not zero, we have poles at infinity,
and can not use the original field to conduct the
integration, since our algorithm can't handle poles at infinity.
So we need to transform into a field where the differential has no pole at infinity.
It's fairly obvious that the original differential has no poles at zero,
since differentials can only have poles at {\it finite} places if their
functions also have poles there, but just to check this:

\begin{maximablock}
puiseux(y^2 - (1 - x^2), x, y, 0, false,
        -1, 1/y * del(x));
\end{maximablock}

See, no poles at zero!  We'll use the change of variable $z=\frac{1}{x}$
to swap zero with infinity to get our differential into a form
that our algorithm can handle:

 $${1\over y}\,dx = {1\over y}\,d(\frac{1}{z}) = - \frac{1}{yz^2}\,dz$$

$$y^2=1-x^2 = 1-\frac{1}{z^2} \qquad \Longrightarrow \qquad y^2 z^2 = z^2 -1$$

So, we're going to integrate $-\frac{1}{yz^2}$ on the algebraic curve
$y^2 z^2=z^2 - 1$, which has no poles at infinity.

Our next step
is to identify all of the points where we can have poles, which is
simply all of the points where the denominator $yz^2=0$.  One
of those points is $z=0$, and plugging $y=0$ into $y^2 z^2=z^2-1$,
we see that this condition is met when $z=\pm i$.

Now let's compute the principle parts at each of these places:

\begin{maximablock}
puiseux(y^2 * z^2 - (z^2 - 1), z, y, 0, false,
        -1, -1/(y*z^2) * del(z));
puiseux(y^2 * z^2 - (z^2 - 1), z, y, %i, false,
        -1, -1/(y*z^2) * del(z));
puiseux(y^2 * z^2 - (z^2 - 1), z, y, -%i, false,
        -1, -1/(y*z^2) * del(z));
\end{maximablock}

The places with zero principle parts can be ignored; what happened
there is that the function had a pole but the differential did not.
We're interested in the other two, with residues $-i$ and $i$.  These
exist in the field ${\bf Q}[i]$, which can be regarded as a vector
field over ${\bf Q}$ with basis $\{1, i\}$, and we want to construct a
function whose poles and zeros match the $i$-component of the residues
(the 1-component is uniformly zero).

So, we want a function
with a simple zero at $(0,i)$ and a simple pole at $(0,-i)$.

We can proceed using the algorithm of the previous chapter to
construct a basis for the Riemman-Roch field $L(D)$, where $D$ is the
divisor with a simple zero at $(0,i)$ and a simple pole at $(0,-i)$.
The basis will have either one element or no elements, depending
on whether an algebraic function exists with the desired properties.

Yes, it does exist, and it's ${{v-i}\over{u}}$.

Now we convert back to our original field:

$${{v-i}\over{u}} = x({y\over x}-i) = y - ix $$

Remembering that our residues came multiplied by a factor of $i$, we
conclude that our solution is $i\,\ln(y-ix)$, or:

\begin{eqnarray*}
\int {1\over{\sqrt{1-x^2}}} \,dx &=& i\,\ln(\sqrt{1-x^2}-ix) \\
                                 &=& -i\,\ln({1\over{\sqrt{1-x^2}-ix}}) \\
                                 &=& -i\,\ln({{\sqrt{1-x^2}+ix}\over{1-x^2+x^2}}) \\
                                 &=& -i\,\ln({\sqrt{1-x^2}+ix}) \\
                                 &=& \arcsin x \\
\end{eqnarray*}

where I used the negative of a logarithm being the logarithm of the
inverse, and the last transformation came from section
\ref{sec:Liouvillian Forms}.


\endexample

\example Compute $\int \sqrt{4-x^2} \,dx$

A solution method from first year calculus might be to note that this
integrand forms one leg of a right triangle with its other sides $2$
and $x$, but we'll attack this integral using the methods of this
chapter.

First, transform the problem into an algebraic curve:

$$\int y\,dx \qquad y^2 = 4-x^2$$

Since $\lim_{x\to\infty} y = \infty$, the integrand has poles at
infinity.  We want infinity to be an ordinary point of the curve (no
ramification; no singularities) with no poles in the integrand.  The
simplest transformation is again to exchange zero with infinity, and in this
case zero is an ordinary point with places $(0,2)$ and $(0,-2)$,
neither of which is a pole of the integrand.  So we'll invert
$x$ and $y$ into $u$ and $v$:

$$x=\frac{1}{u} \qquad y=\frac{1}{v}$$
$$\left(\frac{1}{v}\right)^2 = 4 - \left(\frac{1}{u}\right)^2 \Longrightarrow 4u^2v^2 - v^2 - u^2=0$$
$$\int\frac{1}{v} \, d\left(\frac{1}{u}\right) \Longrightarrow -\int\frac{1}{vu^2}\,du$$

The only poles in this integrand occur when either $u=0$ or $v=0$.
Substituting these values into $4u^2v^2 - v^2 -u^2=0$, we see that
these condiutions only occur at $(u,v)=(0,0)$, so let's analyze our
curve at that point, starting with the Newton polygon:

\begin{center}
$4 u^2 v^2 - v^2 - u^2 = 0$ \\
\setlength{\unitlength}{1cm}
\begin{picture}(3,3)
\put(0,0){\line(0,1){2.5}}
\put(0,0){\line(1,0){3}}
\put(1.9,-0.1){x}
\put(1.9,1.9){x}
\put(-0.1,1.9){x}
\thicklines
\put(0,2){\line(1,-1){2}}
\end{picture}
\end{center}

The Newton polygon has a single line segment of span 2 and slope -1, so
we have two cycles, each with ramification index one: a singularity.
Since there is no ramification, $u$ is a uniformizing parameter
and we expect to expand $v$ as follows:

$$v = c_1 u + c_2 u^2 + c_3 u^3 + \cdots$$
$$v^2 = c_1^2 u^2 + 2 c_1 c_2 u^3 + (2 c_1 c_3 + c_2^2) u^4 + \cdots$$

Substituting these expansions into $4u^2v^2 - v^2 - u^2 = 0$, we obtain:

$$ 4 c_1^2 u^4 + 8 c_1 c_2 u^5 + (8 c_1 c_3 + 4 c_2^2) u^6 + \cdots $$
$$ - c_1^2 u^2 - 2 c_1 c_2 u^3 - (2 c_1 c_3 + c_2^2) u^4 + \cdots - u^2 = 0$$

Equating terms in $u^2$, we see that $c_1 = \pm i$.  Each of these
two values corresponds to one branch of the singularity.  There
is only a single term in $u^3$, which forces $c_2$ to be zero,
and equating terms in $u^4$ produces $c_3 = 2 c_1$, so

$$v = \pm (iu + 2iu^3 + \cdots) \qquad @(0,0)$$

Inverting $v$ and substituting into our 1-form, we obtain

$$\frac{1}{v} = \pm (-i \frac{1}{u} + 2i u + \cdots) \qquad @(0,0)$$

$$\frac{1}{vu^2}\, du = \pm \left[ -i \frac{1}{u^3} + 2i \frac{1}{u} + \cdots \right] \, du \qquad @(0,0)$$

The $u^{-1}$ terms will integrate into logarithms, so let's ignore
them for the moment and concentrate on the $u^{-3}$ terms, which will
integrate into $u^{-2}$ terms, so we're looking for a function with
second order poles at both places at the $(0,0)$ singularity.

Starting with our standard basis for all rational functions,
$\{1,\,v\}$, we seek to modify it into a basis for
${\rm P}^2(0,0)_a{\rm P}^2(0,0)_b$.  Note first that $v$ has
poles at $u=\pm\frac{1}{2}$.  Using $y=1/u$, we analyze
at $(\pm\frac{1}{2}, \infty)$ as follows:

\begin{center}
$y^2\left[(u-\frac12)^2+(u-\frac12)+\frac14\right]-4(u-\frac12)^2-4(u-\frac12)$
\\
\setlength{\unitlength}{1cm}
\begin{picture}(3,3)
\put(0,0){\line(0,1){2.5}}
\put(0,0){\line(1,0){3}}
\put(0.9,-0.1){x}
\put(1.9,-0.1){x}
\put(-0.1,1.9){x}
\put(0.9,1.9){x}
\put(1.9,1.9){x}
\thicklines
\put(0,2){\line(1,-2){1}}
\end{picture}
\end{center}

Our line segment has span 1 and slope -2, indicating a single place
with ramification 2, and $y$ as a uniformizing parameter.  Setting

$$(u-\frac12) = c_1 y + c_2 y^2 + \cdots$$
$$(u-\frac12)^2 = c_1^2 y^2 + \cdots$$

Substituting, we find that $c_1 = 0$ and $c_2 = \frac{1}{16}$, so

$$(u-\frac12) = \frac{1}{16} y^2 + \cdots \qquad v=y^{-1} \qquad @(\frac12, \infty)$$

$$(u+\frac12) = \frac{1}{16} y^2 + \cdots \qquad v=y^{-1} \qquad @(-\frac12, \infty)$$

In short, $v$ has first order poles at $(\pm\frac12,\infty)$ and
$(u\pm\frac12)$ has second order zeros, so we can adjust our basis
accordingly and obtain $\{1,\,(4u^2-1)v\}$ for a basis with no finite
poles.  We can also use a theorem of Trager to shortcut this calculation.

Returning to our analysis at $(0,0)$, we see that 1 has zero order
(obviously) and $(4u^2-1)v$ has a first order zero at both sheets
there, since $4u^2-1=-1$ is finite and $v$ has first order zeros.
We also know that $u$ is a uniformizing parameter, so it's easy
to modify our basis and obtain

$$\left\{\frac{1}{u^2},\,\frac{4u^2-1}{u^3}v\right\} {\rm is\, a\,} {\bf C}[x]{\rm -basis\, for\, P^2(0,0)_aP^2(0,0)_b}$$

Is this basis normal at infinity?  Well, the representation order of
$\frac{1}{u^2}$ is 2 and its $u^-2$ coefficients at $(\infty, \pm
\frac12)$ are both 1, while the representation order of $\frac{4u^2-1}{u^3}v$
is 1, and its $u^-1$ coefficients are 2 and -2.  Since

$$\det C = \begin{array}{|cc|} 1 & 2 \\ 1 & -2 \end{array} = -4$$

is non-zero, the basis is normal at infinity.

The Riemann-Roch theorem says that the dimension of ${\mathfrak l}(D)$ is 5,
$\frac{1}{u^2}$ can be multiplied by any polynomial up to second
degree without introducing poles at infinity, and $\frac{4u^2-1}{u^3}v$
can be multiplied by any polynomial up to first degree, so

$$\left\{\frac{1}{u^2},\, \frac{1}{u},\, 1,\, \frac{4u^2-1}{u^3}v,\, \frac{4u^2-1}{u^2}v\right\}$$

is a ${\cal C}$-module basis for ${\mathfrak l}(D)$.

Any linear combination of these functions is a multiple of the
divisor, but not all of them produce the correct residues.  Looking at
the residues, we see that only $\frac{4u^2-1}{u^3}v = \frac{1}{uv}$
has residues of $\pm i$ on the two sheets at the $(0,0)$ singularity.
Dividing by 2 to correct for the 2 that will be introduced by the
integration, we conclude that $\frac{1}{2uv} = \frac{xy}{2} =
\frac{x\sqrt{4-x^2}}{2}$ is the desired function.

Next, we have to deal with the logarithms.  Going back to the
series expansions of our 1-form, we see that we have residues
of $\pm 2i$ on our two sheets at $(0,0)$.  The objective
now is a bit different; we want a function with exactly
the divisor $Z(0,0)_a P(0,0)_b$.  Starting with an integral basis:

$$\{1, (4u^2-1)v\}$$

we want to modify these functions to make them multiples
of $Z(0,0)_a P(0,0)_b$.  The pole isn't a problem for
an integral basis, and looking at the series expansion
for $v$ at $(0,0)$ we see that it (and therefore $(4u^2-1)v$)
has a simple zero there, but $1$ needs to be replaced with $u$:

$$\{u, (4u^2-1)v\}$$

Now we construct a matrix with the coefficients in the series expansions:

$$\left[ \begin{array}{cc} 1 & -i \\ 0 & 0 \end{array} \right] \begin{array}{ll} \leftarrow (0,0)_a \\ \leftarrow (0,0)_b \end{array} $$

$$\left[ \begin{array}{cc} 1 & -i \\ 0 & 0 \end{array} \right] \left[ \begin{array}{c} i \\ 1 \end{array} \right] = 0$$

The solution shows us how to modify the basis:

$$\{u, \frac{iu + (4u^2-1)v}{u}\} = \{u, i + \frac{(4u^2-1)v}{u}\}$$

$$\left[ \begin{array}{cc} 1 & 0 \\ 0 & 0 \end{array} \right] \begin{array}{ll} \leftarrow (0,0)_a \\ \leftarrow (0,0)_b \end{array} $$

$$\left[ \begin{array}{cc} 1 & 0 \\ 0 & 0 \end{array} \right] \left[ \begin{array}{c} 0 \\ 1 \end{array} \right] = 0$$

$$\{u, i\frac{1}{u} + \frac{(4u^2-1)v}{u^2}\}$$

$$\left| \begin{array}{cc} 1 & -2i \\ 0 & 2i \end{array} \right| = 2i$$

At the last step, the determinant is non-zero, which shows that we
now have a basis for multiples of the divisor except at infinity.
Is it normal at infinity?  $u$'s expansion at both places at infinity
is $\left(\frac{1}{u}\right)^{-1}$, so its representation order is -1,
and the second element's expansion at infinity starts $\pm 2 + \cdots$,
so its representation order is 0 and:

$$\det C = \begin{array}{|cc|} 1 & 2 \\ 1 & -2 \end{array} = -4$$

So the basis is normal at infinity.  If an exact multiple of
the divisor exists, it is one of the basis elements.  It's not $u$,
since $u$ has a pole at infinity, but the second element is exact:

$$i\frac{1}{u} + \frac{(4u^2-1)v}{u^2} = i\frac{1}{u} - \frac{1}{v} = ix-y$$

The desired residues are $\pm 2i$, so the function we want is

$$2i \ln(ix-y) = 2i \ln(\frac{y}{2}-i\frac{x}{2}) + 2i \ln(-2) $$
$$= 2i \ln\left(\sqrt{1-\left(\frac{x}{2}\right)^2} - i\frac{x}{2}\right) = 2i (-i \arcsin \frac{x}{2}) = 2 \arcsin \frac{x}{2}$$

(the constant disappears into the constant of integration) and the final answer is:

$$ \int \sqrt{4-x^2} \, dx  = 2\arcsin\frac{x}{2} + \frac{x \sqrt{4-x^2}}{2}$$

\endexample

\vfill\eject
\mysection{Geddes's example}

\example Compute $\int {1\over{x\sqrt{x^4+1}}} \, dx$

We'll use ${\bf C}(x,y); y^2=x^4+1$ and integrate ${1\over{xy}} =
{y\over{x^5+x}}$.  Inverting this field ($z={1\over x}$) shows that
this integrand has no poles at infinity, so we can proceed directly:

$$ {y\over{x^5+x}} = {y\over{x(x+\omega)(x-\omega)(x+i\omega)(x-i\omega)}} \qquad \omega = \sqrt{i} = {\sqrt{2}\over2} + {\sqrt{2}\over2} i $$

\bigskip
\begin{center}
\begin{supertabular}{l l l}
  $(0, 1)$  &  $\displaystyle {y\over{x^4+1}}$ @ $(0, 1)$     & = $1$    \cr
  $(0, -1)$  &  $\displaystyle {y\over{x^4+1}}$ @ $(0, -1)$     & = $-1$    \cr
  $(\omega, 0)$  &  $\displaystyle 2{y\over{x(x^2+i)(x+\omega)}}$ @ $(\omega, 1)$     & = $0$    \cr
  \multicolumn{2}{l}{$(-\omega, 0), (i\omega, 0), (-i\omega, 0)\qquad\cdots$}    & = $0$    \cr
\end{supertabular}
\end{center}

We now use theorem \ref{simple pole construction} to construct a
function with a simple pole at $(0,-1)$:

$${{f(0,y)}\over{x(y+1)}} = {{y^2-1}\over{x(y+1)}} = {{y-1}\over{x}} $$

This function has a zero at $(0,1)$, but, unfortunately, it is third order,
as can be seen from either L'H\^opital's rule:

$$y^2=x^4+1$$
$$2y\,dy=4x^3\,dx$$
$${{dy}\over{dx}} = 2{x^3\over y}$$

$${{y-1}\over{x}} @ (0,1) = \lim {{dy}\over{dx}} = 2 {{x^3}\over y} = 0$$
$${{y-1}\over{x^2}} @ (0,1) = \lim {1\over{2x}}{{dy}\over{dx}} = {{x^2}\over y} = 0$$
$${{y-1}\over{x^3}} @ (0,1) = \lim {1\over{3x^2}}{{dy}\over{dx}} = {2\over3}{{x}\over y} = 0$$
$${{y-1}\over{x^4}} @ (0,1) = \lim {1\over{4x^3}}{{dy}\over{dx}} = {1\over2}{1\over y} = {1\over2}$$

\vfil\eject

\ldots or from a series expansion of $y$ at (0,1):

$$y^2 = x^4 + 1 $$
$$(y-1)^2 = x^4 - 2(y-1)$$
$$(y-1) = {1\over2}x^4 - {1\over2}(y-1)^2$$
$$(y-1) = c_0 + c_1 x + c_2 x^2 + c_3 x^3 + \cdots$$
$$(y-1)^2 = c_0^2 + (2 c_0 c_1) x + (2 c_0 c_2 + c_1^2) x^2 + (2 c_0 c_3 + 2 c_1 c_2) x^3 + \cdots$$

$$ c_0, c_1, c_2, c_3 = 0 $$
$$ c_4 = {1\over2}$$
$$ c_5, c_6, c_7 = 0 $$
$$ c_8 = -{1\over8} $$

$$ (y-1) = {1\over2} x^4 - {1\over8} x^8 + \cdots$$

$$ {(y-1)\over x} = {1\over2} x^3 - {1\over8} x^7 + \cdots$$

\ldots or from the norm:

$$N({{y-1}\over{x}}) = {{y-1}\over{x}} \cdot {{-y-1}\over{x}} = - {{y^2-1}\over{x^2}} = - {{x^4}\over{x^2}} = - x^2$$

Since we know that the function has a simple pole at $(0,-1)$, so it
must have a third order zero at $(0,1)$ to form a norm with a second
order zero.

We can eliminate the inconvenient zero by adding a constant to the
function, say 1: ${{x+y-1}\over x}$.  We can now use theorem
\ref{simple zero construction} to create a simple zero at
(0,1) by multiplying by $x+y-1$:

$${{x+y-1}\over x} (x+y-1) = {{2x-2}\over x} y + {{x^4+x^2-2x+2}\over x} $$

This function has a simple pole at (0,-1) and a simple zero at (0,1),
but does it have other poles and zeros?  If so, can it be modified to
eliminate them?  To find out, we form the generators of an ${\cal I}$-module:

$$\{ {{2x-2}\over x} y + {{x^4+x^2-2x+2}\over x}, x \}$$

Noting that ${{x^4+x^2}\over{x}} = x(x^2+1)$ and $x^2+1 \in {\cal I}$,
we can simplify this:

$$\{ {{2x-2}\over x} y - {{2x-2}\over x}, x \}$$

\vfill\eject

Using the integral basis $\{1, y \}$, we convert this to a
${\bf C}[x]$-module:

$$\{ {{2x-2}\over x} y - {{2x-2}\over x}, {{2x-2}\over x} (x^4+1) - {{2x-2}\over x} y, x, xy \}$$

and since $(2x-2){{x^4}\over{x}} = x(2x^3-2x^2)$ and
$2x^3-2x^2 \in {\bf C}[x]$, we simplify:

$$\{ {{2x-2}\over x} y - {{2x-2}\over x}, {{2x-2}\over x} - {{2x-2}\over x} y, x, xy \}$$

and write it in matrix form:

% $$\pmatrix{-{{2x-2}\over{x}} & {{2x-2}\over{x}} \cr {{2x-2}\over{x}} & -{{2x-2}\over{x}} \cr x & 0 \cr 0 & x} \pmatrix{1 \cr y}$$

$${1\over x}\begin{pmatrix}-(2x-2) & 2x-2 \cr 2x-2 & -(2x-2) \cr x^2 & 0 \cr 0 & x^2\end{pmatrix} \begin{pmatrix}1 \cr y\end{pmatrix}$$

Elementary row operations\footnote{Read right to left; $R_{i,j,\lambda}$ adds $\lambda$ times row $j$ to row $i$; $R_{i,\alpha}$ multiplies row $i$ by $\alpha$ (a unit)} $R_{4,3,x^2} R_{1,3,(2x-2)} R_{3,4,-1} R_{3,1,{1\over2}(x+1)} R_{2,1,1} $ yield:

$${1\over x}\begin{pmatrix}1 & -1 \cr x^2 & 0\end{pmatrix} \begin{pmatrix}1 \cr y\end{pmatrix} = \begin{pmatrix}{{1-y}\over{x}} \cr x\end{pmatrix} $$

so $\{ {{1-y}\over{x}}, x \} $ forms a generator set for
the ${\bf C}[x]$-module of the finite multiples of $Z(0,1)P(0,-1)$.
We convert to a basis normal at infinity: $\{1, v\} = \{1, {y\over x^2}\}$:

$$\begin{pmatrix}{{1-y}\over{x}} \cr x\end{pmatrix} = \begin{pmatrix}{1\over x} & -x \cr x & 0\end{pmatrix} \begin{pmatrix}1 \cr {y\over x^2}\end{pmatrix}
= \begin{pmatrix}x & \cr & x\end{pmatrix} \begin{pmatrix}{1\over x^2} & -1 \cr 1 & 0\end{pmatrix} \begin{pmatrix}1 \cr {y\over x^2}\end{pmatrix}$$

$$\det_{@ \infty} \begin{pmatrix}{1\over x^2} & -1 \cr 1 & 0\end{pmatrix} = 1$$

so $\{ {{1-y}\over{x}}, x \} $ is normal at infinity.  $x$ clearly has
a pole at infinity, so it can't be the function we're looking for, but
what about ${{1-y}\over{x}}$?  Switching back to $\{u,v\}$
coordinates, we obtain ${{1-y}\over{x}} = {{u^2-v}\over u}$, which has
$\{u,v\}$ poles at both $\{0,1\}$ and $\{0,-1\}$, which translate into
poles at $x=\infty$.  Therefore, no rational function on this
algebraic curve has a simple pole at (0, -1), a simple zero at (0,1),
and no other poles or zeros.

\vfill\eject

So, let's try a double pole at (0,-1) and a double zero at (0,1).  We
can just square our previous generators:
$\{ {{1-y}\over{x}}, x \} $
to obtain: 
$\{ {(1-y)^2\over{x^2}}, x^2 \} = \{ {{1-2y+x^4+1}\over{x^2}}, x^2 \}$
which simplifies to
$\{ {{1-y}\over{x^2}}, x^2 \}$.  We again check for normalcy
at infinity:

$$\begin{pmatrix}
{1\over{x^2}} & -1 \cr
x^2 & 0\end{pmatrix} \begin{pmatrix}1 \cr {y\over{x^2}}\end{pmatrix} = \begin{pmatrix}1 & \cr & x^2\end{pmatrix}\begin{pmatrix}
{1\over{x^2}} & -1 \cr
1 & 0\end{pmatrix}
\begin{pmatrix}1 \cr {y\over{x^2}}\end{pmatrix} $$

$$\det_{@\infty}\begin{pmatrix}
{1\over{x^2}} & -1 \cr
1 & 0\end{pmatrix} = 1$$

So, $\{{{1-y}\over{x^2}}, x^2\}$ is a ${\bf C}[x]$-module, normal at
infinity, containing the finite multiples of $Z^2(0,1)P^2(0,-1)$.
$x^2$ has a pole at infinity, but does ${1-y}\over{x^2}$?
Switching to $x={1\over z}; y={u\over{z^2}}; u^2 = z^4 + 1$ and
Writing it as $z^2(1-{u\over{z^2}}) = z^2 - u$ shows that it has no
zero at $(z,u) = (0, \pm 1)$, and thus no pole at $x = \infty$.  It
is, therefore, the function we are looking for:

$$\int {1\over{x\sqrt{x^4+1}}} \, dx
   = {1\over2} \ln{{1-\sqrt{x^4+1}}\over{x^2}}$$

I'll now point out to you what's been pointed out to me, and that is a
traditional solution technique for this integral:

$$\int {1\over{x\sqrt{x^4+1}}} \, dx$$
$$x^2 = u \qquad 2 x\, dx = du$$
$$\int {1\over{2u\sqrt{u^2+1}}} \, du$$
$$u = \tan z \qquad du = \sec^2 z dz$$
$${1 \over 2} \int \csc z \, dz = {1\over 2} \ln \tan {z\over2}$$
$$={1\over2} \ln {{\sec z - 1}\over{\tan z}}$$
$$={1\over2} \ln {{\sqrt{u^2+1}-1}\over{u}}$$
$$={1\over2} \ln {{\sqrt{x^4+1}-1}\over{x^2}}$$

\vskip 0.5in

$${{1 - \cos z}\over{\sin z}} = {{1 - \cos^2 {z\over2} + \sin^2 {z\over2}}\over{2\sin{z\over2}\cos{z\over2}}} = {{2\sin^2 {z\over2}}\over{2\sin{z\over2}\cos{z\over2}}} = {{\sin {z\over2}}\over{\cos{z\over2}}} $$

\vfill\eject

We can also proceed like this:

$$\int {1\over{x\sqrt{x^4+1}}} \, dx$$
$$x^4 = u \qquad 4 x^3\, dx = du$$
$$\int {1\over{4u\sqrt{u+1}}} \, du$$
$$v = u + 1 \qquad dv = du$$
$$\int {1\over{4(v-1)\sqrt{v}}} \, dv$$
$$z^2 = v \qquad 2 z dz = dv$$
$$\int {1\over{4(z^2-1)z}} \, 2z\, dz$$
$$\int {1\over{2(z^2-1)}} \, dz$$
$${1\over{z^2-1}} = {1\over2}{1\over{z-1}} - {1\over2}{1\over{z+1}}$$
$${1\over4} \int {1\over{z-1}} - {1\over{z+1}} \, dz$$
$${1\over4} \ln (z-1) - \ln (z+1)$$
$${1\over4} \ln {{z-1}\over{z+1}}$$
$${1\over4} \ln {{\sqrt{v}-1}\over{\sqrt{v}+1}}$$
$${1\over4} \ln {{\sqrt{u+1}-1}\over{\sqrt{u+1}+1}}$$
$${1\over4} \ln {{\sqrt{x^4+1}-1}\over{\sqrt{x^4+1}+1}}$$

so from the previous page and this one, we conclude

$$\int {1\over{x\sqrt{x^4+1}}} \, dx
={1\over2} \ln {{\sqrt{x^4+1}-1}\over{x^2}}
={1\over4} \ln {{\sqrt{x^4+1}-1}\over{\sqrt{x^4+1}+1}}$$

Is this last equality true?  Well, $\ln f^2 = 2\ln f$, so
${1\over4}\ln f^2 = {1\over2}\ln f$, and\ldots

$$\Big({{\sqrt{x^4+1}-1}\over{x^2}}\Big)^2
= {{(\sqrt{x^4+1}-1)^2}\over{x^4}}$$

$${{\sqrt{x^4+1}-1}\over{\sqrt{x^4+1}+1}} \cdot
{{\sqrt{x^4+1}-1}\over{\sqrt{x^4+1}-1}}
= {{(\sqrt{x^4+1}-1)^2}\over{x^4+1-1}}$$




\endexample

\vfill\eject
\mysection{Chebyshev's Integral}

\example Compute:
\label{Chebyshev's Integral}
$$\int {{2x^6+4x^5+7x^4-3x^3-x^2-8x-8}\over{(2x^2-1)^2\sqrt{x^4+4 x^3+2 x^2+1}}} \,{\rm d}x$$

The polynomial under the square root is square-free:

\begin{maximablock}
num : 2*x^6 + 4*x^5 + 7*x^4-3*x^3-x^2-8*x-8;
den: 2*x^2-1;
root : x^4+4*x^3+2*x^2+1;
factor(root);
\end{maximablock}

\ldots so $y^2 = x^4+4 x^3+2 x^2+1$; $\{1, y\}$ is an integral basis;
and our normal form for this integral is:

$$\int {{(2x^6+4x^5+7x^4-3x^3-x^2-8x-8)y}\over{(2x^2-1)^2(x^4+4 x^3+2 x^2+1)}} \,{\rm d}x$$

Applying now Bronstein's Hermite reduction from
section 2.1 of his ``Symbolic Integration Tutorial'' with $v=2x^2-1$
to eliminate this square in the denominator:

\begin{maximablock}
gradef(y,x,ratsimp(D(root)/2*y/root));
D(y);
U: root;
V: den;
S2: ratsimp(U*V^2*D(y/V));
\end{maximablock}

Now we want to solve $f_2 S_2 = A_2 y$ where $A_2 y$ is our numerator.

\begin{maximablock}
kill(f);
f[2] ::: num*y/S2;
T[2] ::: num(f[2]);
/* Q::: S2/y; */
Q ::: denom(f[2]);

[A,R,g] : gcdex(V,Q,x);
/* T2: num; */
[Q2, B2]: divide(T[2]*R, V, x);
h ::: A*num*y/(V*U) - (D(V)*Q2+D(B2))*y/V + Q2*D(y);

simp: false$
'integrate(num*y/den,x) = ratsimp(B2*y/V) + 'integrate(h,x);
simp: true$
\end{maximablock}

\vfill\eject
\bigskip
\begin{center}
Non-zero residues

\begin{supertabular}{r @{} l | r @{} l | r @{} l}
\multicolumn{2}{c|}{x} & \multicolumn{2}{c|}{y} & \multicolumn{2}{c}{residue} \cr
\hline
&$\sqrt{2}\over 2$ & &${1\over 2} + \sqrt{2}$ & &${5\over2}$ \cr
&$\sqrt{2}\over 2$ & $-$&${1\over 2} - \sqrt{2}$ & $-$&${5\over2}$ \cr
$-$&${\sqrt{2}\over 2}$ & &${1\over 2} - \sqrt{2}$ & &${5\over2}$ \cr
$-$&${\sqrt{2}\over 2}$ & $-$&${1\over 2} + \sqrt{2}$ & $-$&${5\over2}$ \cr
\end{supertabular}
\end{center}


$$A(x) = 1023x^8+4104x^7+5048x^6+2182x^5+805x^4+624x^3+10x^2+28x$$
$$B(x) = 1025x^{10} + 6138x^9 + 12307x^8 + 10188x^7 + 4503x^6 + 3134x^5 + 1598x^4 + 140x^3 + 176x^2 +2$$
$$C(x) = 32x^{10}-80x^8+80x^6-40x^4+10x^2-1$$

$$y = \sqrt{x^4+4 x^3+2 x^2+1}$$

$$\int {{(2x^6+4x^5+7x^4-3x^3-x^2-8x-8)}\over{(2x^2-1)^2\sqrt{x^4+4 x^3+2 x^2+1}}} \,{\rm d}x
= {{(x+{1\over2})y}\over{2x^2-1}} + {1\over2}\ln{{A(x)y - B(x)}\over{C(x)}}
$$


\endexample

\vfill\eject
\mysection{The Risch Theorem: A First Look}

At this point, there is only one major missing piece in our
integration theory for Abelian integrals --- how do we limit the
multiples of a divisor to a testable set?  We've seen how to
repeatedly raise a divisor to higher and higher powers, but how do we
know when to stop?  At what point can we declare that a divisor has no
multiple that is principle?

We'll attack this problem the way Robert Risch discovered in 1970, by
mapping into a finite field, solving the corresponding problem there,
then lifting the result back to the original field.  The finite field
will be the integers modulo a prime.  Reducing modulo some primes
changes the genus of the algebraic curve, while other primes leave the
genus unchanged.  Those primes at which the genus of the curve remains
unchanged yield {\it good reduction}

The key theorems as stated by [Tr73] on page 67:

\theorem
The homomorphism between divisor class groups under good reduction
is an isomorphism when restricted to divisors whose orders are
relatively prime to the characteristic of the reduced function field.
\endtheorem

Let $p$ be the characteristic of the reduced field,

\theorem
if the divisor $D$ has order $p^k n$ where $\gcd(n,p)=1$, then
the reduction of $D$ must have order $p^j n$ for some $j\le k$.

\proof
Let the order of the reduction be $p^j m$.  Since reduction
is a group homomorphism, we must have $m|n$ and $j \le k$.
Since $D^{p^k}$ has order exactly $n$, its reduction
must have order exactly $n$.  But the order of its
reduction is a divisor of $m$ and thus $n|m$ and so
finally we have $n=m$.

\endtheorem


Since good reduction preserves the part of the divisor's order
relatively prime to the characteristic, by picking two different
primes we can completely reconstruct the divisor's order in
characteristic zero.

A property of algebraic curves over finite fields is that {\it all}
divisors of total degree zero have some multiple that is principal
(PROOF).  Thus, to carry out this program, we need to compute bases
for Riemann-Roch spaces in prime characteristic.  Then, we can keep
raising a divisor to higher and higher powers until we find its order.
Do this for two different primes, each exhibiting good reduction, and
then we can find the order of the original divisor in characteristic
zero.

To proof the above claims, I want to construct the Jacobian variety in
arbitrary characteristic, then show how Jacobians map under good
reduction.

\section{Simple Algebraic Extensions over Finite Fields}

Let's start with a simple but crucial observation:

\theorem

In an algebraic extension over a finite field, the evaluation field is
also finite.

\proof

Consider a finite field of constants ${\cal F}$, over which we'll
extend first into a rational function field ${\cal F}(x)$ and then add
an algebraic extension ${\cal F}(x,y)$, where $y$ satisfies some
minimial polynomial $f(x,y)=0$.  Start with the constant field, which
gives us a finite number of values for $x$.  Plugging each of these
values into the minimal polynomial gives a finite set of polynomials
$f(y_i)=0$.  By Theorem ?, we can extend ${\cal F}$ into a finite
extension field ${\cal F}[\gamma]$ where all the roots of the
polynomial exist.  Since there a only a finite number of polynomials,
we need at worst a finite set of extensions ${\cal
F}[\gamma_1,...,\gamma_k]$ to construct a field in which all the roots
of all the polynomials exist.  Using the Theorem of the Primitive
Element, we can collapse all of these into a single finite extension
field ${\cal F}[\phi]$.  Since all values of $x$ exist in ${\cal F}$,
and all values of $y$ exist in ${\cal F}[\phi]$, an evaluation
homomorphism carries any rational function in $x$ and $y$ into
${\cal F}[\phi]$.

\endtheorem

This theorem leads directly to the single more important difference
(to us) between divisors in an infinite field versus those in a finite
field.  {\it In a finite field, some multiple of every divisor is
principle.}  The reason is that the multiplicative group of the
evaluation field has finite order.  The simplest way to demonstrate
this is to construct theorems analogous to Theorems ? and ?:

\theorem

In an algebraic extension of a finite field with characteristic
greater than 2, a function can always be constructed with an $m^{\rm
th}$-order zero at a specified place $(\alpha, \beta)$ and zero order
at all other finite places, where $m$ is the multiplicative order of
the evaluation field.

\proof

The desired function is

$$(x-\alpha)^m + (y-\beta)^m$$.

Clearly, this function is zero at $(\alpha, \beta)$ and of $m^{\rm
th}$ order there (PROOF THIS).  At all other places one of the two
terms will be non-zero, and both exist in the evaluation field.  By
Theorem ?, any non-zero number raised to the multiplicative order of
its field is one.  Thus the value of this function will be either
$1+0$, $0+1$, or $1+1=2$, all finite and non-zero, and thus of zero
order.

\endtheorem

\theorem

In an algebraic extension of a finite field with characteristic
greater than 2, a function can always be constructed with an $m^{\rm
th}$-order pole at a specified place $(\alpha, \beta)$ and zero order
at all other finite places, where $m$ is the multiplicative order of
the evaluation field.

\proof

The desired function is

$${f(\alpha,y)^m\over(x-\alpha)^m(y-\beta)^m} + 1$$

where the division by $(y-\beta)^m$ is exact.
Clearly, this function has a pole at $(\alpha, \beta)$ and of $m^{\rm
th}$ order there (PROOF THIS).  CONSIDER OTHER PLACES OVER $\alpha$.
At all other places the denominator
term will be non-zero, and thus one, and the numerator will be
either zero or one (by Theorem ?)
Thus the value of this function at these places will be either
$0+1$ or $1+1=2$, both finite and non-zero, and thus of zero
order.

\endtheorem


\example

Show that some multiple of ${\mathrm Z}(1,1)$ is principle in
${\bf Z}_5(x,y); y^2=x$.

Let's first construct a multiplication table for ${\bf Z}_5$:

\begin{center}
\begin{tabular}{c|c c c c c}
  & 0 & 1 & 2 & 3 & 4 \cr
\hline
0 & 0 & 0 & 0 & 0 & 0 \cr
1 & 0 & 1 & 2 & 3 & 4 \cr
2 & 0 & 2 & 4 & 1 & 3 \cr
3 & 0 & 3 & 1 & 4 & 2 \cr
4 & 0 & 4 & 3 & 2 & 1 \cr
\end{tabular}
\end{center}

Now, let's list out the places on the Riemann surface for
${\bf Z}_5(x,y); y^2=x$.

\begin{center}
\begin{tabular}{c l}
$x$ & $(x,y)$ \cr
\hline
0 & (0,0) \cr
1 & (1,1) \quad (1,4) \cr
2 & $(2,\gamma) \quad (2,-\gamma); \quad \gamma^2 - 2 =0$ \cr
3 & $(3,\theta) \quad (3,-\theta); \quad \theta^2 - 3 =0$ \cr
4 & (4,2) \quad (4,3) \cr
\end{tabular}
\end{center}

It looks like we need ${\bf Z}_5[\gamma,\theta]$ to express these places.
It's simplest to collapse $\gamma$ and $\theta$ into a single algebraic
extension.  We could use the Theorem of the Primitive Element to
do this, but in this case just looking at the multiplication table
and noting that $3 = 2^3 = \gamma^6$ shows that $\theta = \pm \gamma^3$.
So, in fact, we only need ${\bf Z}_5[\gamma]$:

\begin{center}
\begin{tabular}{c l}
$x$ & $(x,y)$ \cr
\hline
0 & (0,0) \cr
1 & (1,1) \quad (1,4) \cr
2 & $(2,\gamma) \quad (2,-\gamma); \quad \gamma^2 - 2 =0$ \cr
3 & $(3,\gamma^3) \quad (3,-\gamma^3)$ \cr
4 & (4,2) \quad (4,3) \cr
\end{tabular}
\end{center}

Since ${\bf Z}_5[\gamma]$ has $5^2=25$ elements, its multiplicative
group has order one less than this.  We conclude that 24 is our
``magic'' multiple, and that ${\mathrm Z}^{24}(1,1)$ must be
principle in this field.  Its generator should be simply
$(x-1)^{24} + (y-1)^{24}$.  Clearly this function is zero for
$(x,y)=(1,1)$.  Let's verify that it's non-zero for some other
places on the Riemann surface:

\begin{eqnarray*}
(0,0) &:& (-1)^{24} + (-1)^{24} = 4^{24} + 4^{24} = 1+1 = 2 \cr
(1,4) &:& 3^{24} + 0^{24} = 1 + 0 = 1 \cr
(2,\gamma) &:& (\gamma-1)^{24} + (2-1)^{24} = 1+1 = 2 {\rm ,\quad since:} \cr
&&\cr
&& (\gamma-1)^2 = (\gamma^2-2\gamma+1) = 3-2\gamma \cr
&& (\gamma-1)^4 = (3-2\gamma)^2 = (9-12\gamma+4\gamma^2) = 2-2\gamma \cr
&& (\gamma-1)^8 = (2-2\gamma)^2 = (4-8\gamma+4\gamma^2) = 2-3\gamma \cr
&& (\gamma-1)^{12} = (2-2\gamma)(2-3\gamma) = (4-10\gamma+6\gamma^2) = 1 \cr
\end{eqnarray*}

In the final series of calculations, I used $\gamma^2=2$ and reduced
mod 5 repeatedly.  I think the pattern should be clear, and leave
further verification as an exercise.

\endexample


