
% \mychapter{Bibliography}

\begin{thebibliography}{9}

\bibitem[Al14]{alvanos}
Paraskevas Alvanos;
{\it Riemann-Roch Spaces and Computation}.
De Gruyter Open, 2014.
ISBN 978-3-11-042613-7, e-ISBN 978-3-11-042612-0

Free PDF download available from publisher's web site

\bibitem[Bl47]{bliss}
Gilbert Ames Bliss; {\it Algebraic Functions}. American Mathematical Society, 1947
ISBN 0821874535, 9780821874530

\bibitem[Br91]{bronstein algebraic curve}
Bronstein, Manuel; {\it The Risch Differential Equation on an Algebraic Curve}. 1991.

\bibitem[Br00]{bronstein tutorial}
Bronstein, Manuel; {\it Symbolic Integration Tutorial}. 2000.

\bibitem[Br05]{bronstein book}
Bronstein, Manuel; {\it Symbolic Integration I}. Springer, 2005.

\bibitem[BrCo10]{briggs}
Briggs, Cochran, {\it Calculus}, $2^{nd}$ Edition.  Pearson, 2010.  ISBN-10: 0321336119

\bibitem[DeWe82]{dedekind-weber}
Dedekind, R., Weber, H. (1882). Theorie der algebraischen Functionen einer Veanderlichen.
{\it J. reine angew. Math.}, {\bf 92}:181-290.

\bibitem[Fu08]{fulton}
Fulton, William, {\it Algebraic Curves, An Introduction to Algebraic Geometry}, $3^{rd}$ Edition, 2008.

{\tt http://www.math.lsa.umich.edu/$\sim$wfulton/CurveBook.pdf}

\bibitem[Ge92]{geddes}
Geddes, Czapor, Labahn, {\it Algorithms for Computer Algebra}. Springer.
ISBN: 978-0-585-33247-5

\bibitem[Go14]{goodman}
Goodman, {\it Algebra: Abstract and Concrete}, $6^{th}$ Edition.  SemiSimple Press.
ISBN: 978-0-9799142-1-8

\bibitem[Gu05]{guillemin}
Victor Guillemin. 18.117 Topics in Several Complex Variables. Spring 2005. Massachusetts Institute of Technology: MIT OpenCourseWare, https://ocw.mit.edu. License: Creative Commons BY-NC-SA.

\bibitem[He02]{hess}
Hess, F. (2002). Computing Riemann-Roch Spaces in Algebraic Function Fields and Related Topics,
Journal of Symbolic Computation, 33(4), 425-445.

\bibitem[Ko07]{kollar}
Kollár, János (2007), Lectures on Resolution of Singularities, Princeton: Princeton University Press, ISBN 978-0-691-12923-5

Kollár's Resolution of Singularities -- Seattle Lecture, {\tt https://arxiv.org/abs/math/0508332},
is a more advanced treatment of resolution of singularities on varieties in general, while
his Princeton University Press book contains two introductory chapters on resolution for curves
(the only one relevant here) and resolution for surfaces before it gets into the more general theory.

\bibitem[Sh61]{shimura}
Shimura, Goro; Taniyama, Yutaka (1961), {\it Complex multiplication of abelian varieties and its applications to number theory},
Publications of the Mathematical Society of Japan, 6, Tokyo: The Mathematical Society of Japan, MR 0125113

Later expanded and published as Shimura (1997)

\bibitem[SwHu06]{Swanson Huneke}
Swanson, Irena; Huneke, Craig (2006), {\it Integral Closure of Ideals, Rings, and Modules},
Cambridge University Press, Cambridge.

Available on-line at {\tt https://www.math.purdue.edu/~iswanso/book/}

\bibitem[Ri70]{risch}
Robert Risch; {\it The Solution of the Problem of Integration in Finite Terms}.  Bulletin A.M.S, Vol 76, pp. 605-608.

\bibitem[Ru76]{baby rudin}
Rudin, Walter (1976). {\it Principles of mathematical analysis (Third ed.)}. New York. ISBN 0-07-054235-X.

\bibitem[Tr84]{trager}
Trager, Barry; {\it Integration of Algebraic Functions}, Ph.D. thesis, Massachusetts Institute of Technology, 1984.

% How to Compute a Puiseux Expansion
% https://arxiv.org/abs/0807.4674

\end{thebibliography}
